Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-08 paper/main.tex
some proofing, filling some methods gaps

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the annotation of 
genes in eukaryotic genomes have to be pretrained for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for those tools also remains the same since about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA data alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-are tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

\section{Introduction}
The annotation of genes in DNA is fundamental for many other tasks depending on it 
to be done as good and correct as possible. It can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, gene annotation is a very 
labour intensive process and genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl} offer gene 
annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homolog sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State of the art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble to generalize across species and it seems 
actual learning of underlying patterns is limited despite the fundamental ability 
to do so being present.

@@

For the purpose of gene annotation, RNNs have already shown promising results. 
\citep{choudharypredicting} carries out preliminary explorations on the potential 
of BLSTMs for cross species gene prediction and trains his model on human genes to 
test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses 
bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in 
prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes 
are proportionately more frequent in the genome, feature simpler control structures 
and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes 
the use of a BLSTM after a concurrent neural net (CNN) to find detailed motifs in 
the human genome. DeePromotor \citep{oubounyt2019deepromoter} trains a similar 
architecture for the recognition of promotor regions. Recently, several groups 
\citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to 
find splicing sites.

In this work we present Helixer, a novel prototype software for training and 
utilizing a general purpose DNN for the ab-initio cross-species gene annotation of 
large eukaryotic genomes using only DNA data as input. Our model is trained to 
differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and 
Intron. We demonstrate the effectiveness of this approach by training a single model 
for the annotation of a large set of genomes from the kingdoms metazoa and 
viridiplantae, respectively, which we will call {\it animals} and {\it plants} from 
now on. We worked with the full data of 192 animal genomes and 60 plant genomes. 
Both the ability to generalize across genomes as well as the scope of the evaluation 
greatly surpasses all other work in this field. The source code and all input data 
is publically available.

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. We used all animals genomes 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant 
genomes of the JGI Phytosome 13 database \citep{goodstein2012phytozome}. Both data 
groups were used seperately during training as well as during evaluation.

At the start of our work we split both genome groups into {\it training genomes} and 
{\it test genomes}. The test genome groups are made up of 19 and 6 genomes, 
respectively, and were chosen to be of the best possible quality based on collected 
metadata while also representing a broad phylogenetic spread.

The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we do want to train with as much data as possible, but 
not all data has a good enough quality to enable a powerful generalization. It is 
also very desirable to have a broad spread phylogenetically as well as in genome 
size and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, if the genomic patterns 
inside the data are too different from each other, it may be difficult for the model 
to generalize, too. It was also important for us to get experimental results within 
a couple of days as we tested many different data and model configurations, which 
limited the overall size of the training set and the model.

We employed an iterative approach to effectively select a proper set of training 
genomes. We first started with 3-4 genomes that were almost certain to be of the 
highest quality. We then evaluated the performance of the resulting model on all 
genomes individually (including the training genomes) and looked for candidates to 
add to the training set or to remove from it. This process was repeated multiple 
times concurrently with the model search until we saw no more room for substantial 
improvement given our computational constraints.

>>>>

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
% normally contact is last author / person with permanent position (i.e. Andreas). 
% Not sure how important that is, but if we're skipping that, then we can change it back to Felix
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

\section{Introduction}
The annotation of genes in DNA is fundamental for many downstream tasks, which 
depend on high quality annotations. Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{}, PASA \citep{} or those offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and it seems 
actual learning of underlying patterns is limited despite the fundamental ability 
to do so being present \citep{}.

@@

For the purpose of gene annotation, RNNs have already shown promising results. 
\citep{choudharypredicting} carries out preliminary explorations on the potential 
of BLSTMs for cross species gene prediction and trains his model on human genes to 
test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses 
bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in 
prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes 
are proportionately more frequent in the genome, feature simpler control structures 
and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes 
the use of a BLSTM after a convolutional neural net (CNN) to find detailed motifs in 
the human genome. DeePromotor \citep{oubounyt2019deepromoter} trains a similar 
architecture for the recognition of promotor regions. Recently, several groups 
\citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to 
find splicing sites.

In this work we present Helixer, a novel prototype software for training and 
utilizing a general purpose DNN for the ab-initio cross-species gene annotation of 
large eukaryotic genomes using only DNA sequence as input. Our model is trained to 
differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and 
Intron. We demonstrate the effectiveness of this approach by training a single model 
for the annotation of a large set of genomes from the kingdoms metazoa and 
viridiplantae, respectively, which we will call {\it animals} and {\it plants} from 
now on. We worked with the full data of 192 animal genomes and 60 plant genomes. 
Both the ability to generalize across genomes as well as the scope of the evaluation 
greatly surpasses all other work in this field. The source code and all input data 
is publically available.

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant
species from the JGI Phytosome 13 database \citep{goodstein2012phytozome}; exact 
genomes used are list here \ref{}. % add a supplemental table with names and versions
Data was downloaded on 29/03/2019 and 15/10/2019 for plants and animals respectively. 
Both data groups were used seperately during training as well as during evaluation.

Both to divide genomes into training, validation and test sets, as well as to understand
performance, we needed a way to evaluate genome and annotation quality.
Metadata was generated for each genome, including: basic length and composition statistics \citep{}; % quast
k-mer content \citep{}; % jellyfish
conserved gene content of the genome, transcriptome, and proteome \citep{}; % BUSCO, files generated with gffread
and finally simple counts of genic features from the gff files. In particular, 
we considered the following features of the metadata when judging the quality of a genome and it's 
annotation: assembly contiguity (N50 / genome size); retention in the proteome 
of detected conserved orthogroups (BUSCOs) that were identified in the genome; and
the presence of UTRs and alternative splicing as estimated by the gff feature-counts.
Finally this metadata was combined with researcher knowledge of model species to guide genome
selection as necessary.

At the start of our work we split both genome groups into {\it training genomes} and 
{\it test genomes}. The test genome groups are made up of 19 and 6 genomes, 
respectively, and were chosen to be of decent quality and diversity based on collected 
metadata while also representing a broad phylogenetic spread.

For both feasibility and iterative cross-species generalization validattion, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we do want to train with as much data as possible, but 
not all data has a good enough quality to enable a powerful generalization. It is 
also very desirable to have a broad spread phylogenetically as well as in genome 
size and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, if the genomic patterns 
inside the data are too different from each other, it may be difficult for the model 
to generalize, too. It was also important for us to get experimental results within 
a couple of days as we tested many different data and model configurations, which 
limited the overall size of the training set and the model.

We employed an iterative approach to effectively select a proper set of training 
genomes. We first started with 3-4 genomes that were expected to be of the 
highest quality. We then evaluated the performance of the resulting model on all 
genomes individually (including the training genomes) and looked for candidates to 
add to the training set or to remove from it. This process was repeated multiple 
times concurrently with the model search until we saw no more room for substantial 
improvement given our computational constraints.
